---
title: "Data Analysis with R and Claude Code"
subtitle: "Day 1: Foundations — Why Data Analysis & First Steps in R"
author: "Fabio Giglietto"
institute: "University of Urbino"
date: "February 9, 2026"
format:
  revealjs:
    theme: [default, uniurb.scss]
    slide-number: true
    preview-links: auto
    logo: ""
    footer: "PhD in Humanities — Scienze della Comunicazione e Cultura Digitale"
    highlight-style: github
    code-line-numbers: false
    code-overflow: wrap
    scrollable: true
    margin: 0.05
execute:
  echo: true
  eval: false
---

# Welcome! {background-color="#2c3e50"}

## About This Course

::: {.incremental}
- **5 days** (Mon-Fri, 2 hours each)
- **No prior programming experience required**
- Learn by doing with a real dataset
- AI-assisted learning with Claude Code
:::

. . .

**Goal:** Go from zero to analyzing social media data

## About Me

**Fabio Giglietto**

- University of Urbino
- Research interests: social media, misinformation, computational methods

. . .

**About You?**

- Your name
- Your research area
- What do you hope to learn?

## This Week's Journey {.smaller}

| Day | Topic |
|-----|-------|
| **1** | **Foundations** — First steps in R |
| 2 | Working with Data — Import & explore |
| 3 | Data Wrangling — Transform & summarize |
| 4 | Visualization — Tell stories with data |
| 5 | Text Analysis — Capstone project |

# Why Data Analysis? {background-color="#2c3e50"}

## The Media Research Landscape

::: {.columns}
::: {.column width="50%"}
**Traditional Methods**

- Interviews
- Focus groups
- Content analysis (manual)
- Small samples
:::

::: {.column width="50%"}
**Computational Methods**

- Large-scale data collection
- Automated analysis
- Pattern detection
- Reproducible research
:::
:::

## Mixed Methods

Data analysis **complements** qualitative research

. . .

- **Quantitative:** What patterns exist?
- **Qualitative:** Why do they exist?

. . .

Best insights often come from combining both

## Questions Data Can Help Answer

::: {.incremental}
- How do stories spread on social media?
- What content gets engagement?
- How do audiences respond to events?
- Who are the key voices in a conversation?
- What patterns exist in large text collections?
:::

# Our Dataset {background-color="#2c3e50"}

## TikTok Videos about Fabrizio Corona

We'll analyze **9,373 TikTok videos** collected in January 2026

. . .

| Attribute | Value |
|-----------|-------|
| Videos | 9,373 |
| Unique creators | 4,964 |
| Total views | 338+ million |
| Period | January 2026 |
| Region | Italy |

## How Was This Data Collected?

Collected via the **TikTok Research API** using two search criteria:

::: {.incremental}
1. **Keyword**: videos with "Fabrizio Corona" in the description
2. **Hashtag**: videos tagged with `#fabriziocorona`
:::

. . .

Filtered by region (Italy) and date (January 2026). Duplicates removed.

## A Note on Data Quality

This is a **real-world dataset** — it's messy!

::: {.incremental}
- **False positives**: some videos may mention Corona without being truly "about" him
- **Missing data**: not all videos have descriptions or hashtags
- **Snapshot in time**: engagement counts were captured on Feb 4, 2026
:::

. . .

**This is normal.** Cleaning and understanding data is part of the process.

# Getting Started {background-color="#2c3e50"}

## Project Setup Check

You should have:

::: {.incremental}
1. Downloaded the course ZIP from GitHub
2. Extracted it to a folder (e.g., `Documents/r-course`)
3. Created an RStudio Project (**File → New Project → Existing Directory**)
:::

. . .

Check: **Project name in top-right corner** of RStudio

## If You Need to Catch Up

1. Download: [course GitHub repository](https://github.com/fabiogiglietto/r-data-analysis-uniurb-2026) → **Code** → **Download ZIP**
2. Extract the ZIP to a folder
3. In RStudio: **File → New Project → Existing Directory**
4. Browse to the extracted folder → **Create Project**

## Your Project Structure {.smaller}

In the **Files** pane (bottom right), you should see:

```
├── data/           ← Our TikTok dataset
├── scripts/        ← R scripts for each day
├── resources/      ← Stopwords, cheatsheets
├── output/         ← Your saved plots
└── slides/         ← Slide decks
```

. . .

**Why this matters:** file paths like `"data/fabrizio_corona_tiktok_jan2026.csv"` will just work!

# RStudio Orientation {background-color="#2c3e50"}

## The RStudio Interface

:::: {.columns}
::: {.column width="50%"}
**Top Left: Source**

- Write and save code
- Your R scripts live here
:::

::: {.column width="50%"}
**Top Right: Environment**

- Variables you've created
- Data you've loaded
:::
::::

:::: {.columns}
::: {.column width="50%"}
**Bottom Left: Console**

- Run code interactively
- See output here
:::

::: {.column width="50%"}
**Bottom Right: Files/Plots/Help**

- Browse files
- View visualizations
- Read documentation
:::
::::

## Writing vs Running Code

**Source pane:** Write code to save and reuse

**Console:** Run code immediately, see results

. . .

**Key shortcut:** `Ctrl+Enter` (or `Cmd+Enter` on Mac) — runs the current line

## Using the Daily Scripts

Open today's script: **File → Open File → scripts/day1-basics.R**

. . .

**How to follow along:**

::: {.incremental}
- The script mirrors these slides — follow along section by section
- Run each line with `Ctrl+Enter` (`Cmd+Enter` on Mac)
- Complete exercises marked `# [YOUR CODE HERE]`
- Add your own notes and experiments
:::

. . .

The scripts are your **workbook** for this course!

## Try It: Your First Code

Type in the Console:

```{r}
2 + 2
```

. . .

Press Enter. You should see:

```
[1] 4
```

# R Basics {background-color="#2c3e50"}

## R as a Calculator

```{r}
2 + 2        # Addition
10 - 3       # Subtraction
5 * 4        # Multiplication
20 / 5       # Division
3^2          # Exponentiation (3 squared)
sqrt(16)     # Square root
```

## Try It Yourself

Calculate:

::: {.incremental}
1. 15 + 27
2. 100 divided by 8
3. 5 to the power of 3
4. Square root of 144
:::

## Variables: Storing Values

Use `<-` to assign values to variables

```{r}
x <- 10
y <- 5
```

. . .

Now use them:

```{r}
x + y
x * y
```

## Naming Variables

```{r}
my_name <- "Fabio"        # Text (character)
my_age <- 30              # Number (numeric)
is_student <- TRUE        # Logical (TRUE/FALSE)
```

. . .

View a variable by typing its name:

```{r}
my_name
```

```
[1] "Fabio"
```

## Data Types

| Type | Example | Description |
|------|---------|-------------|
| `numeric` | `42`, `3.14` | Numbers |
| `character` | `"hello"` | Text (in quotes) |
| `logical` | `TRUE`, `FALSE` | Boolean values |

## Vectors: Collections of Values

Use `c()` to **combine** values:

```{r}
views <- c(1000, 5000, 25000, 100000, 50000)
views
```

```
[1]   1000   5000  25000 100000  50000
```

## Vectors of Different Types

```{r}
# Numeric
views <- c(1000, 5000, 25000, 100000, 50000)

# Character
platforms <- c("TikTok", "Instagram", "YouTube")

# Logical
is_viral <- c(TRUE, FALSE, TRUE, TRUE, FALSE)
```

## Accessing Vector Elements

Use square brackets `[ ]`:

```{r}
views <- c(1000, 5000, 25000, 100000, 50000)

views[1]       # First element
views[3]       # Third element
views[c(1,3)]  # First AND third
```

```
[1] 1000
[1] 25000
[1]  1000 25000
```

## Useful Vector Functions

```{r}
views <- c(1000, 5000, 25000, 100000, 50000)

length(views)   # How many elements?
sum(views)      # Total
mean(views)     # Average
max(views)      # Maximum
min(views)      # Minimum
```

## Sorting

```{r}
views <- c(1000, 5000, 25000, 100000, 50000)

sort(views)                      # Ascending
sort(views, decreasing = TRUE)   # Descending
```

```
[1]   1000   5000  25000  50000 100000
[1] 100000  50000  25000   5000   1000
```

## Logical Operations

Comparisons return `TRUE` or `FALSE`:

```{r}
5 > 3    # Greater than
5 < 3    # Less than
5 == 5   # Equal to (two equals signs!)
5 != 3   # Not equal to
```

## Applying Comparisons to Vectors

```{r}
views <- c(1000, 5000, 25000, 100000, 50000)

views > 10000   # Which are greater than 10000?
```

```
[1] FALSE FALSE  TRUE  TRUE  TRUE
```

. . .

```{r}
sum(views > 10000)   # How many?
```

```
[1] 3
```

## Getting Help

Use `?` before any function:

```{r}
?mean
?sum
?sort
```

. . .

The Help pane shows:

- Description
- Usage
- Arguments
- Examples

# Introduction to Claude Code {background-color="#2c3e50"}

## What is Claude Code?

An AI assistant that helps you:

::: {.incremental}
- **Explain** code you don't understand
- **Write** code for specific tasks
- **Debug** errors
- **Learn** programming concepts
:::

## Starting Claude Code

1. Click the **Terminal** tab (next to Console)
2. Type `claude` and press Enter

. . .

Claude Code can see your files and help with your project!

## Example: Explain Code

Ask Claude Code:

> "What does this code do?"
>
> `views <- c(1000, 5000, 25000)`

. . .

Claude will explain:

- `c()` combines values into a vector
- `<-` assigns the vector to a variable called `views`

## Example: Debug Errors

Ask Claude Code:

> "I got this error: object 'Views' not found"
>
> "My code is: mean(Views)"

. . .

Claude will explain: R is case-sensitive!

You probably defined `views` (lowercase).

## Best Practices with AI

::: {.incremental}
1. **Try first** — Attempt problems yourself
2. **Ask for explanations** — Not just code
3. **Understand** — Make sure you can explain it
4. **Practice** — Try similar problems without AI
:::

. . .

**Learn WITH AI, not FROM AI**

# Exercises {background-color="#2c3e50"}

## Exercise 1

Create variables for:

- Your name (as text)
- Your birth year (as a number)
- Whether you like coffee (TRUE or FALSE)

## Exercise 2

Create a vector called `likes` with values: 100, 500, 2000, 800, 350

Then calculate:

1. The mean
2. The sum
3. The maximum value
4. How many values are above 500?

## Exercise 3

Use Claude Code!

Ask it: "What is the difference between `<-` and `=` in R?"

# Homework {background-color="#2c3e50"}

## Before Day 2

**1. Verify tidyverse is loaded**

```{r}
library(tidyverse)
```

You should see a list of packages. If you get an error, run `install.packages("tidyverse")`.

. . .

**2. Practice in RStudio**

Complete the exercises in `scripts/day1-basics.R` if you haven't already.

## Ask Claude Code

> "What is the tidyverse and why is it useful for data analysis?"

. . .

Read the response and come ready to discuss tomorrow!

# See You Tomorrow! {background-color="#2c3e50"}

**Day 2: Working with Data**

- Import our TikTok dataset
- Explore and understand the data
- Select and filter observations
